# Age-Structured Models

## The recipe
### Ingredients
For more details on data needs, see XXX, but in brief:

1. A list of annual survival rates (the fraction of individuals that survive for a year), one for each age. 
Ideally, this will be survival from one birthay to the next (including from birth to the first birthday). 
If males and females have different survival rates, list the female values. 
We will call this list the vector $\vect{p}$, with elements $p_0$ (survival probability from birth to age 1), $p_1$ (survival probability from age 1 to age 2), $p_2$ (survival probability from age 2 to age 3), etc.
2. A list of age-specific birth rates: For each parental age, the number of offspring per parent (counting both parents), or the number of female offspring produced by each female parent. 
We will call this list the vector $\vect{b}$, with elements $b_1$ (newborns produced by a 1-year-old parent), $b_2$ (newborns produced by a 2-year-old parent), etc. 
Newborns cannot reproduce, so there is no value of $b_0$; if the age at first reproduction is less than one, then you will need a shorter timestep.
3. Identify the maximum age, above which there is no survival. 
Call this age $m$; while there may be a non-zero birth rate $b_m$, $m$-year-olds do not survive to age $m+1$, so $p_m = 0$. 

When these are arranged in tabular form, it will look like part of a life table, which you may recall from undergraduate ecology courses.
In fact, these two lists provide all the information needed to construct the other columns of the life table (such as the cumulative survivorship, $l_x$); but we don't need those to construct an MPM, so don't pursue that further.


```{r lizard-image,echo=FALSE, fig.cap="The viviparous lizard (*Zootoca vivipara*) in the Aamsveen, The Netherlands. CC-BY-SA [Ocrdu](https://commons.wikimedia.org/wiki/User:Ocrdu)."}
knitr::include_graphics("figs/Viviparous_lizard_(Zootoca_vivipara)_in_the_Aamsveen,_The_Netherlands.jpg")
```

The example we will use in this chapter comes from @Strijbosch1988, who quantified the age-specific demography of the viviparous lizard (*Zootoca vivipara*; Fig. \@ref(fig:lizard-image)) in an intensive 7-year study on a 1.2 ha plot at the forest reserve "de Hamert" in the Netherlands.
The age-specific female birth and survival rates are shown in Table \@ref(tab:lizard-lt).

```{r lizard-lt, echo=FALSE}
px <- c(0.424, 308/424, 158/308, 57/158, 10/57, 0.7, 2/7) # Survival 
bx <- c(0.08, 2.94, 4.13, 4.88, 6.5, 6.5, 6.5)/2          # Births
lizard <- data.frame(x = 0:7, px = c(round(px, 3), 0), bx = c(NA, bx))
knitr::kable(
  lizard,
  caption = "Age-specific survival and birth rates for the lizard _Zootoca vivipara_ (formerly _Lacerta vivipara_) in the Netherlands [@Strijbosch1988]. Age-specific annual survival is calculated from the female cumulative survival values ($l_x$) in their Table 6. Age-specific birth rates are calculated by dividing the birth rates in their Table 6 ($m_x$) by two, as these appear to represent all offspring, and the primary sex ratio (estimated from two-year-old individuals) did not differ from 1 (pp. 21-22).",
  col.names = c("Age in years ($x$)", "Female annual survival ($p_x$)", "Female offpring per female ($b_x$)") 
)
```

### Construct the birth matrix, $\vect{B}$
1. Make an $m \times m$ identity matrix (with ones on the diagonal, and zeros everywhere else)
2. Add an extra row on top, with the birth rates

If $m=3$, this would look like:
\begin{equation}
\vect{B} = \begin{pmatrix}
              b_1 & b_2 & b_3 \\
              1 & 0 & 0 \\
              0 & 1 & 0 \\
              0 & 0 & 1
            \end{pmatrix}
(\#eq:B3-age)
\end{equation}

### Construct the survival matrix, $\vect{P}$
1. Make an $m \times m$ matrix with the survival values on the diagonal, and zeros everywhere else
2. Add an extra column of zeros on the right

If $m=3$, this would look like:
\begin{equation}
\vect{P} = \begin{pmatrix}
              p_0 & 0 & 0 & 0 \\
              0 & p_1 & 0 & 0 \\
              0 & 0 & p_2 & 0
            \end{pmatrix}
\end{equation}

### Combine the birth and survival matrices into the projection matrix
You have two choices here: 

* the *postbreeding census model* (which we denote $\vect{A_{post}}$), which takes the population immediately after the breeding season in one yeqr (including the newborns) and projects it to immediately after the breeding season the next year;  
* the *prebreeding census model* (which we denote $\vect{A_{pre}}$), which takes the population immediately before the breeding season in one yeqr (so that the youngest individuals are just about to reach their first birthday) and projects it to immediately before the breeding season the next year.

Both versions model the same population; the only difference is when you peek in to see the population abundances. Here are the formulas:

\begin{equation}
\begin{aligned}
\vect{A_{post}} &= \vect{B} \vect{P}\\
\vect{A_{pre}} &= \vect{P} \vect{B}
\end{aligned}
\end{equation}

The formulas look similar, differing only in the order of multiplication. But remember that matrix multiplication is not commutative! 

For the $m=3$ case above, these work out to:
\begin{equation}
\begin{aligned}
\vect{A_{post}} &= \begin{pmatrix}
              p_0 b_1 & p_1 b_2 & p_2 b_3 & 0\\
              p_0 & 0 & 0 & 0 \\
              0 & p_1 & 0 & 0 \\
              0 & 0 & p_2 & 0
            \end{pmatrix} \\
\vect{A_{pre}} &= \begin{pmatrix}
              b_1 p_0 & b_2 p_0 & b_3 p_0 \\
              p_1 & 0 & 0 \\
              0 & p_2 & 0
            \end{pmatrix} 
\end{aligned}
(\#eq:A3-pre-post)
\end{equation}

## Example
Let's apply this to the lizard life table in Table \@ref(tab:lizard-lt). This model is large enough that I don't want to calculate it by hand, so I'll use R. 
First, enter the age-specific survival and birth rates, and the maximum age:
```{r}
px <- c(0.424, 0.726, 0.513, 0.361, 0.175, 0.700, 0.286) # Survival 
bx <- c(0.040, 1.470, 2.065, 2.440, 3.250, 3.250, 3.250) # Births
m  <- 7                                                  # Maximum age
```
Notice that the first element of `px` is $p_0$, whereas the first element of `bx` is $b_1$.
Thus, `px` doesn't have the zero surivival of 7-year-olds, and `bx` doesn't have the zero birth rate of newborns.
If you are working with a dataset that includes these zeros, you'll need to trim them off before moving to the next step.

Now construct the birth matrix, $\vect{B}$.
A useful function in R is `diag()`, the diagonal function.
If `m` is a scalar (a single number), then `diag(m)` creates an $m \times m$ matrix with ones on the diagonal and zeros everywhere else.
Also, recall that `rbind()` ("row bind") adds the second argument below the first argument.
```{r}
B <- diag(m) # Make an m x m diagonal matrix
B <- rbind(bx, B) # Add an extra row on top, with the birth rates
B
```
ALthough it's larger, this matrix has the same structure as Eq. (\@ref(eq:B3-age)).

One quirk of R is that, while the columns are labeled with the index of the columns (which is what we expect for a generic matrix), the first row is labeled "bx" and the rest of the rows are unlabeled.
The labels don't actually matter for the rest of the calculations, but they will result in the final matrices having uninformative row labels.
If you don't like this, you can reset the row labels with the command `rownames(B) <- NULL`.
Alternatively, you can use this slightly more complex command (it also does both steps at once):
```{r}
B <- matrix(c(bx, diag(m)), 
            nrow = m + 1, 
            ncol = m, 
            byrow = TRUE)
B
```
Now construct the survival matrix, $\vect{P}$.
We will again use `diag()`, but in a different way.
If `px` is a vector with length $m$, then `diag(px)` creates an $m \times m$ matrix with the values of `px` on the diagonal and zeros everywhere else.
Also, recall that `cbind()` ("column bind") adds the second argument to the right the first argument.
```{r}
P <- diag(px)     # Make the diagonal matrix with survival values
P <- cbind(P, 0)  # Add an extra column of zeros
P
```

Finally, we use matrix multiplication to get the projection matrices.
In R, matrix multiplication is indicated with the compound symbol `%*%` (if you just used `*` then it would do element-by-element multiplication, which is mathematically very different).
```{r}
# Make the postbreeding census model
Apost <- B %*% P
Apost

# Make the prebreeding census model
Apre <- P %*% B
Apre
```
If you work through the numbers, you'll see that these match the structure of the post-breeding and pre-breeding matrices in Eq. \@ref(eq:A3-pre-post).

In theory, these two matrices should have exactly the same asymptotic growth rate, $\lambda$.
Lets use the `lambda()` function in the **popbio** package to check this:
```{r}
library(popbio)
lambda(Apost)
lambda(Apre)
```
Yup, they match!

## Why use this recipe?
Now, these matrices don't look that complex; why didn't I tell you how to construct them directly, rather than constructing the intermediate birth and survival matrices? Look at the matrix elements on the top row of each matrix; these are called the *fertility coefficients*, and represent the reproducive component of the model. Look at how they are constructed from survival and birth rates: does this make sense to you? If it does not, don't worry: it is not an implementation of biological intuition, but instead a bunch of bookkeeping. But first I would need to convince you that these are the right formulas; and then I would need to look over your shoulder and remind you to use them each time you constructed a model, because, like double-entry bookkeeping, it is easy to forget what to do unless you do it often.

## Extensions and caveats
### No maximum age
The recipe and example above describes a classic Leslie matrix model, in which there is a well-defined maximum age at which all individuals die.
In many species, however, individuals may survive more-or-less indefinitely; above a certain age, it might be biologically reasonable to assume that all indiviudals have the same birth and survival rates, independent of age.
Even if there is not strong evidence to support that assumption, there may be so few old individuals in the population that it is impossible to estimate separate demographic rates for each age class.
In this case, we can augment the Leslie model by adding an "old individual" stage; while this technically becomes a stage-structured model, it is such a common modification to age structured models that we include it here.

Now, instead of $m$ being the maximum age, we can write $m+$ to represent the class of individuals age $m$ and above. 
They have a non-zero survival, so the vector of survival values now has $m+1$ values ($p_0, p_1, \dots, p_{m-1}, p_{m+}$).
The only modification to the recipe above is to put the value of $p_{m+}$ into the bottom-right element of the $\vect{P}$ matrix.
All else proceeds as above.

### Post-reproductive ages
In many animal species, individuals may survive beyond their maximum age of reproduction.
These post-reproductive age classes have birth rates of zero, and the resulting projection matrix may be problematic for certain sorts of analyses.
There is no problem with using the matrix to numerically simulate the age-structured population through time.
However, analyses that require calculating the eigenvalues and eigenvectors of the matrix, such as the asymptotic population growth rate, $\lambda$, and the sensitivity of $\lambda$ to matrix elements, might generate mysterious errors (depending on the software being used).
If this occurs, you can simply delete the rows and columns of the projection matrix that are associated with the post-reproductive ages, and re-do the analysis.
Because population growth is unaffected (at least in simple density-independent models such as these) by the survival of post-reproductive ages, you can set those sensitivity values to zero after analyzing the remainder of the matrix.

### Extended breeding season

### Dependent offspring

### Individual heterogeneity

## Sensitivity analyses
This book is about MPM construction, not the analysis of such models. 
Nevertheless, the construction of the annual projection matrix by way of separate birth and survival matrices creates a way to easily calculate the sensitivity and elasticity of the asymptotic growth rate (usually denoted $\lambda$) to changes in the age-specific birth and death rates. 
Existing tools (e.g., the `sensitivity()` function in the **popbio** R package) can calculate sensitivities of $\lambda$ to changes in the elements of an annual projection matrix ($\vect{A_{post}}$ or $\vect{A_{pre}}$). 
However, the matrix entries may be functions of multiple vital rates, and individual vital rates may appear in multiple matrix entries (see eqs. \@ref(eq:A3-pre-post)).
Extracting the sensitivities of $\lambda$ to vital rates from the sensitivities of $\lambda$ to matrix elements requires careful accounting, as well as the application of basic differential calculus; doing this by hand is tedious and error prone.

Instead, we can make use of established procedures for sensitivy analysis of "periodic matrix products" [@Caswell1994].
The recipe is as follows:

1. Construct both the postbreeding census and prebreeding census projection matrices ($\vect{A_{post}}$ and $\vect{A_{pre}}$).
2. Use a standard toolkit to constuct the sensitivity matrices of the two projection matrices.
We will call those $\vect{S_A}^{\text{(post)}}$ and $\vect{S_A}^{\text{(pre)}}$.
3. Construct the *transpose* of the $\vect{B}$ and $\vect{P}$ matrices by swapping the rows and columns. 
We will call these $\vect{B^T}$ and $\vect{P^T}$.
For the $m=3$ example, these look like 
\begin{equation}
\vect{B^T} = \begin{pmatrix}
              b_1 & 1 & 0 & 0\\
              b_2 & 0 & 1 & 0\\
              b_3 & 0 & 0 & 1
            \end{pmatrix}
\end{equation}
and
\begin{equation}
\vect{P^T} = \begin{pmatrix}
              p_0 & 0 & 0\\
              0 & p_1 & 0\\
              0 & 0 & p_1\\
              0 & 0 & 0
            \end{pmatrix}.
\end{equation}
4. Matrix multiply $\vect{P^T}$ by $\vect{S_A}^{\text{(pre)}}$. The top row in the resulting matrix contains the sensitivities of $\lambda$ to the age-specific birth rates.
5. Matrix multiply $\vect{B^T}$ by $\vect{S_A}^{\text{(post)}}$. The diagonal entries in the resulting matrix contains the sensitivities of $\lambda$ to the age-specific survival rates.



### Example
```{r}
library(popbio) # This library includes tools for calculating lambda and matrix sensitivities

# Confirm that our two A matrices give the same asymptotic growth rate
lambda(Apost)
lambda(Apre)

# Calculate the sensitivity matrices
Spost <- sensitivity(Apost)
Spre <- sensitivity(Apre)

# Do the matrix multiplication to calculate the sensitivities to survival
Stemp1 <- t(B) %*% Spost
diag(Stemp1[, 1:m])

# Do the matrix multiplication to calculate the sensitivities to birth rates
Stemp2 <- t(P) %*% Spre
Stemp2[1, ]
```